---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## [🚧 WIP 🚧] `jive`

<!-- badges: start -->
<!-- badges: end -->

The goal of jive is to implement jackknife instrumental-variable estimators (JIVE) and various alternatives. 

## Installation

You can install the development version of jive like so:

``` r
devtools::install_github("kylebutts/jive") 
```

## Example Usage

We are going to use the data from Stevenson (2018). Stevenson leverages the quasi-random assignment of 8 judges (magistrates) in Philadelphia to study the effects pretrial detention on several outcomes, including whether or not a defendant subsequently pleads guilty. 
```{r}
library(jive)
data(stevenson)
```

### Juke n' JIVE

```{r}
jive(
  guilt ~ i(black) + i(white) | bailDate | jail3 ~ 0 | judge_pre,
  data = stevenson
)
```

```{r}
ujive(
  guilt ~ i(black) + i(white) | bailDate | jail3 ~ 0 | judge_pre,
  data = stevenson
)
```

```{r}
ijive(
  guilt ~ i(black) + i(white) | bailDate | jail3 ~ 0 | judge_pre,
  data = stevenson
)
```


### (Leave-out) Leniency Measures 

The package will allow you to estimate (leave-out) leniency measures: 

```{r}
# TODO :-) 
# I have the function, just thinking of the API
```

## Econometric Details on JIVE, UJIVE, IJIVE, and CJIVE

Consider the following instrumental variables setup

$$
  y_i = T_i \beta + W_i' \psi + \varepsilon_i,
$$

where $T_i$ is a scalar endogenous variable, $W_i$ is a vector of exogenous covariates, and $\varepsilon_i$ is an error term. Additionally, assume there is a set of valid instruments, $Z_i$. The intuition of two-stage least squares is to use the instruments to "predict" $T_i$:

$$
  T_i = Z_i' \pi + W_i \gamma + \eta_i.
$$ 

Then, the prediction, $T_i$, is used in place of $T_i$ in the original regression. 

When the dimension of $Z_i$ grows with the number of observations, two-stage least squares is biased (Kolesar, 2013). Without getting into the details, the problem comes because you're predicting $T_i$ using $i$'s own observation in the first-stage. For this reason, Angrist, Imbens, and Krueger (1999) developed the jackknife instrumental-variable estimator (JIVE). In short, for each $i$, $T_i$ is predicted using the first-stage equation leaving out $i$'s own observation. 

In general, the JIVE estimator (and variants) are given by

$$
  \frac{P' Y}{P' T}
$$

where $P$ is a function of $W$ and $T_i$. The particulars differ across the JIVE, the unbiased JIVE (UJIVE), the improved JIVE (IJIVE), and the cluster JIVE (CJIVE). 

### JIVE definition
**Source:** Kolesar (2013) and Angrist, Imbens, and Kreuger (1999)

The original JIVE estimate produces $T$ given by

$$
  T_{JIVE} = (I - D_{(Z,W)})^{-1} (H_{(Z,W)} - D_{(Z,W)}) T,
$$

where $H_{(Z,W)}$ is the hat/projection matrix for $(Z,W)$ and $D_{(Z,W)}$ is the diagonal matrix with the diagonal elements from $H_{Z, W}$.

$$
  P_{JIVE} = M_W T_{JIVE}
$$

### UJIVE definition 
**Source:** Kolesar (2013) 

$$
  T_{UJIVE} = (I - D_{(Z,W)})^{-1} (H_{(Z,W)} - D_{(Z,W)}) T = T_{JIVE}
$$

$$
  P_{UJIVE} = T_{UJIVE} - (I - D_{W})^{-1} (H_{W} - D_{W}) T
$$

### IJIVE definition
**Source:** Ackerberg and Devereux (2009)

Start residualizing $T$, $Y$, and $Z$ by the covariates $W$. The definition they give (assuming things have already been residualized) is

$$
  T_{IJIVE} = P_{IJIVE} = (I - D_{\tilde{Z}})^{-1} (H_{\tilde{Z}} - D_{\tilde{Z}}) \tilde{T}
$$

Note that $P = T$ because you have already residualized by $W$. 


### CJIVE definition
**Source:** Frandsen, Leslie, and McIntyre (2023)

This is a modified version of IJIVE as proposed by Frandsen, Leslie, and McIntyre (2023). This is necessary if the errors are correlated within clusters (e.g. court cases assigned on the same day to the same judge). The modified version is given by:

$$
  T_{CJIVE} = P_{CJIVE} = (I - \mathbb{D}(P_Z, \{ n_1, \dots, n_G \}))^{-1} (H_{\tilde{Z}} - D_{\tilde{Z}}) \tilde{T},
$$

where $\mathbb{D}(P_Z, \{ n_1, \dots, n_G \})$ is a block-diagonal matrix equal to the projection matrix of $Z$ zerod out except for each cluster.

In this package, the same adjustment (replacing the diagonal $D$ with the cluster block-diagonal $\mathbb{D}$ version) can be done for all three estimators, though the details have only been worked out for IJIVE in particular (free research idea).

### Standard errors 

Heteroskedastic-robust standard errors are given by 

$$ 
  \frac{\sqrt{\sum_i P_i^2 \hat{\epsilon}_i^2}}{\sum_i P_i T_i},
$$

where $\hat{\epsilon}_i = M_W Y_i - M_W T_i * \hat{\beta}$. See https://github.com/kolesarm/ivreg/blob/master/ivreg.pdf for derivations.

### Advantages of alternative estimators:

Quoting from the papers that propose each: 

- UJIVE: "UJIVE is consistent for a convex combination of local average treatment effects under many instrument asymptotics that also allow for many covariates and heteroscedasticity"
- IJIVE: "We introduce two simple new variants of the jackknife instrumental variables (JIVE) estimator for overidentified linear models and show that they are superior to the existing JIVE estimator, significantly improving on its small-sample-bias properties"
- CJIVE: "In settings where inference must be clustered, however, the [IJIVE] fails to eliminate the many-instruments bias. We propose a cluster-jackknife approach in which first-stage predicted values for each observation are constructed from a regression that leaves out the observation's entire cluster, not just the observation itself. The cluster-jackknife instrumental variables estimator (CJIVE) eliminates many-instruments bias, and consistently estimates causal effects in the traditional linear model and local average treatment effects in the heterogeneous treatment effects framework."

## Sources

[Ackerberg and Deverux (2009)](literature/Ackerberg_Deverux_2009.pdf)

[Angrist, Imbens, and Krueger (1999)](literature/Angrist_Imbens_Krueger_1999.pdf)

[Frandsen, Leslie, and McIntyre (2023)](literature/Frandsen_Leslie_McIntyre_2023.pdf)

[Kolesar (2013)](literature/Kolesar_2013.pdf)

[kolesarm/ivreg](https://github.com/kolesarm/ivreg)

[ghpk-metrics/stata-manyiv](https://github.com/gphk-metrics/stata-manyiv)
